Visualization and analysis of probability distributions of large temporal data

### 2021-11-08 Rmd/01-intro.Rmd
changed chapter 1 intro

  Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes classes of time deconstructions using linear and cyclic time granularities. Linear granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic granularities can be circular such as hour-of-the-day, quasi-circular such as day-of-the-month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering: hour-of-the-day and second-of-the-minute are single-order-up. Hour-of-the-week is multiple-order-up, because it passes over day-of-the-week. Methods are provided for creating all possible granularities for a time index. A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together (a "harmony"), or when they cannot (a "clash").

   Time granularities can be used to create data visualizations to explore for periodicities, associations and anomalies. The granularities form categorical variables (ordered or unordered) which induce groupings of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables.

   The methods implemented in the open source R package `gravitas` are consistent with a tidy workflow, with probability distributions examined using the range of graphics available in `ggplot2`.

 <!-- This chapter contains a literature review related to linear granularities. -->

@@

This work is a natural extension of Chapter \@ref(ch:gravitas). Many, many other displays might be built using different granularities, such as day-of-week, day-of-month, weekday/weekend, and so on. However, only a handful of them may be interesting, revealing major patterns. Determining which displays exhibit "significant" distributional differences between cyclic granularity categories and plotting only these would allow for more efficient exploration. Furthermore, a few of the displays in this collection will be more engaging than others. This chapter provides a new distance metric for selecting and ranking the multiple granularities. The statistical significance of potential visual discoveries is aided by selecting a threshold for the proposed numerical distance measure. The distance measure is computed for a single or pairs of cyclic granularities, and it can be compared across different cyclic granularities as well as a collection of time series. This chapter also includes a case study using residential smart meter data from Melbourne to demonstrate how the suggested methodology may be utilized to automatically find temporal granularities with significant distributional differences. The methods are implemented in the open-source R package 'hakear.'

@@

In this chapter, we look at the problem of using clustering to discover patterns in a large number of univariate time series across multiple temporal granularities. Time series clustering research is gaining traction as more data is collected at finer temporal resolution, over longer time periods, and for a larger number of individuals/entities. Many disciplines have noisy, patchy, uneven, and asynchronous time series that make it difficult to search for similarities. We propose a method for overcoming these constraints by calculating distances between time series based on probability distributions at various temporal granularities. Because they are based on probability distributions, these distances are resistant to missing or noisy data and aid in dimension reduction. When fed into a clustering algorithm, the distances can be used to divide large data sets into small pockets of similar repetitive behavior. These subgroups can then be analyzed separately or used as distinct prototype behaviors in classification problems. The proposed method was tested on a group of residential electricity consumers from the Australian smart meter data set to show that it can generate meaningful clusters. This chapter includes a brief review of the literature on traditional time series clustering and, more specifically, clustering residential smart meter data.

@@

The thesis is structured as follows. Chapter \@ref(ch:gravitas) provides details of the cyclic granularities, different classes, and computation, and also its usage in exploratory time series analysis through applications. This is implemented in the R package **gravitas**. Chapter \@ref(ch:hakear) provides guidance on how to choose significant cyclic granularities, which are likely to have interesting patterns across its categories. This is available as the R package **hakear**. The chapter @ref (ch: gracsr) provides similarity measures for comparing multiple time series.This is in the developing R package **gracsr**. Chapter \@ref(ch:conclusion) summarizes the software tools developed for the work, and discusses some future plans.

>>>>

Chapter \@ref(ch:gravitas) describes classes of time deconstructions using linear and cyclic time granularities, which can be used to create data visualizations to explore periodicities, associations, and anomalies. It provides a formal characterization of cyclic granularities which can be circular (such as hour-of-the-day), quasi-circular (such as day-of-the-month), and aperiodic (public holidays). It further facilitates manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra, as well as provides a recommendation algorithm to check the feasibility of creating plots for any two cyclic granularities. Our proposed method is also applicable to non-temporal hierarchical granularities with an underlying ordered index. The methods are implemented in the open-source R package `gravitas` and are consistent with a tidy workflow [@wickham2016r], with probability distributions examined using the range of graphics available in `ggplot2`[@Wickham2009pk].

@@

This work is a natural extension of Chapter \@ref(ch:gravitas). Many, many displays might be built using cyclic granularities. However, only a handful of them may be interesting, revealing major patterns. Determining which displays exhibit "significant" distributional differences between cyclic granularity categories and plotting only these would allow for more efficient exploration. Furthermore, a few of the displays in this collection will be more engaging than others. Chapter \@ref(ch:hakear) provides a new distance metric for selecting and ranking the multiple granularities. The statistical significance of potential visual discoveries is aided by selecting a threshold for the proposed numerical distance measure. The distance measure is computed for a single or pairs of cyclic granularities, and it can be compared across different cyclic granularities as well as a collection of time series. This chapter also includes a case study using residential smart meter data from Melbourne to demonstrate how the suggested methodology may be utilized to automatically find temporal granularities with significant distributional differences. The methods are implemented in the open-source R package 'hakear.'

@@

In Chapter \@ref(ch:gracsr), we look at the problem of using clustering to discover patterns in a large number of univariate time series across multiple temporal granularities. Time series clustering research is gaining traction as more data is collected at finer temporal resolution, over longer time periods, and for a larger number of individuals/entities. Many disciplines have noisy, patchy, uneven, and asynchronous time series that make it difficult to search for similarities. We propose a method for overcoming these constraints by calculating distances between time series based on probability distributions at various temporal granularities. Because they are based on probability distributions, these distances are resistant to missing or noisy data and aid in dimension reduction. When fed into a clustering algorithm, the distances can be used to divide large data sets into small pockets of similar repetitive behavior. These subgroups can then be analyzed separately or used as distinct prototype behaviors in classification problems. The proposed method was tested on a group of residential electricity consumers from the Australian smart meter data set to show that it can generate meaningful clusters. This chapter includes a brief review of the literature on traditional time series clustering and, more specifically, clustering residential smart meter data.

@@

The thesis is structured as follows. Chapter \@ref(ch:gravitas) provides details of the cyclic granularities, different classes, and computation, and also its usage in exploratory time series analysis through applications. This is implemented in the R package **gravitas**. Chapter \@ref(ch:hakear) provides guidance on how to choose significant cyclic granularities, which are likely to have interesting patterns across its categories. This is available as the R package **hakear**. The chapter \@ref(ch:gracsr) provides methods to explore heterogeneity in repetitive behavior for multiple time series over multiple cyclic granularities. This is in the developing R package **gracsr**. Chapter \@ref(ch:conclusion) summarizes the software tools developed for the work, and discusses some future plans.
